---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.11.4
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# X dosage compensation

```{r}
source(here::here("scripts/init.R"))
```

We start by looking at the CNA data

```{r}
options(repr.plot.width = 8, repr.plot.height = 4)
p1 <- cna %>% ggplot(aes(x=cna_val)) + geom_density() + xlim(0,5) + geom_vline(xintercept=c(0,0.5, 1.5, 2.5, 3.5, 4.5), color="red") 
p2 <- cna %>% ggplot(aes(x=cna_val, color=factor(cna_round))) + geom_density() + xlim(0,5)  + guides(color=FALSE)
p1 + p2
```

## X analysis


We then go to extract loci that are associated with xist expression

```{r}
xist_loci <- get_xist_loci()

xist_loci_meth <- get_xist_loci_meth()
xist_meth <- xist_loci_meth %>%
    group_by(samp) %>%
    summarise(meth = mean(meth, na.rm = TRUE))

xist_expr <- get_gene_expression_mat() %>%
    filter(name == "XIST") %>%
    gather("samp", "expr", -(chrom:name3.chr))

xist_loci_expr <- get_xist_loci_expr()
df <- xist_meth %>%
    left_join(xist_expr %>% select(samp, expr)) %>%
    left_join(samp_data %>% select(samp, ER = ER1)) %>%
    filter(!is.na(ER))

head(xist_loci)
count(xist_loci, chrom)
```

We show that Xist expression is correlated with XIST loci methylation

```{r}
head(df)
```

#### Extended Data Figure 9B

```{r}
options(repr.plot.width = 7, repr.plot.height = 7)
df %>% group_by(ER) %>% summarise(cor = cor(meth, expr, use="pairwise.complete.obs"))
p_xist_meth_cor <- df %>%
    arrange(sample(samp)) %>% 
    ggplot(aes(x = meth, y = expr, color = ER)) +
    geom_point(size=0.2) + 
    scale_color_manual(values = annot_colors$ER1, guide = FALSE) +
    theme(aspect.ratio = 1) +
    xlab("Methylation") +
    ylab("Xist expression") +
    annotate("text", x = 0.45, y = 6, label = sprintf("~ rho == %0.2f", cor(df$meth, df$expr, method = "spearman", use = "pairwise.complete.obs")), parse = TRUE, size = 2, family = "Arial")
p_xist_meth_cor
```

Moving on to look at the methylation in different CNA regimes

```{r}
get_xist_cna
```

```{r}
xist_cna <- get_xist_cna()
xist_cna %>% count(cna) %>% ggplot(aes(x=factor(cna), y=n)) + geom_col() + theme_bw()
```

Plotting the distribution of loci


#### Extended Data Figure 9D

```{r}
options(repr.plot.width = 8, repr.plot.height = 5)
df <- xist_cna %>%
    left_join(get_xist_loci_meth()) %>%
    group_by(cna_grp, ER, chrom, start, end) %>%
    summarise(meth = mean(meth, na.rm = TRUE)) %>%
    filter(!is.na(ER)) %>% 
    mutate(cna_grp = factor(cna_grp, levels = c("1N", "2N", ">=3N"))) %>% 
    mutate(ER = factor(ER, levels = c("ER+", "ER-")))
p_boxp_meth_cna <- df %>% 
    ggplot(aes(x = cna_grp, y = meth, fill = ER, group = cna_grp)) +
#     ggrastr::geom_boxplot_jitter(outlier.size = 0.1, outlier.jitter.width = 0.01, raster=TRUE) +
    geom_boxplot(linewidth=0.1, fatten=0.5, outlier.size=0.1) + 
    scale_fill_manual(values = annot_colors$ER1, guide = FALSE) +
    xlab("") +
    ylab("Methylation in Xist\nassociated promoters") +
    facet_grid(. ~ ER) +
    ylim(0, 1.1) + 
            theme(
                panel.grid.major.x = element_blank(),
                panel.grid.minor.x = element_blank()
            )
p_boxp_meth_cna  + ggpubr::stat_compare_means(label = "p.signif", comparisons = list(c("1N", "2N"), c("2N", ">=3N")))+ theme_bw()
p_boxp_meth_cna  + ggpubr::stat_compare_means(comparisons = list(c("1N", "2N"), c("2N", ">=3N")))+ theme_bw()
```

```{r}
df %>% ungroup() %>% distinct(chrom, start, end, ER, cna_grp) %>% count(ER, cna_grp)
```

```{r}
xist_cna %>% distinct(cna_grp, samp, ER) %>% count(cna_grp, ER) 
```

We now move to show scatters of loci in different CNA regimes

```{r}
get_xist_meth_cna
xist_meth_cna <- get_xist_meth_cna()
head(xist_meth_cna)
```

#### Extended Data Figure 9E

```{r}
p_scatter_meth_cna_2_3 <- xist_meth_cna %>%
        ggplot(aes(x = `2N`, y = `>=3N`, color = ER)) +
#         ggrastr::geom_point_rast(size = 0.2, raster.width = 1, raster.height = 1) +
        geom_point(size=0.2) + 
        geom_abline(color = "black", linetype = "dashed") +
        scale_color_manual(values = annot_colors$ER1, guide = FALSE) +
        theme(aspect.ratio = 1) +
        xlim(0, 0.65) +
        ylim(0, 0.65) #+
#         annotate("text", x = 0.5, y = 0.1, label = paste("rho = ", round(cor(xist_meth_cna$`2N`, xist_meth_cna$`>=3N`, use = "pairwise.complete.obs"), digits = 3)), size = 2, family = "Arial")
p_scatter_meth_cna_1_2 <- xist_meth_cna %>%
        ggplot(aes(x = `1N`, y = `2N`, color = ER)) +
#         ggrastr::geom_point_rast(size = 0.2, raster.width = 1, raster.height = 1) +
        geom_point(size=0.2) + 
        geom_abline(color = "black", linetype = "dashed") +
        scale_color_manual(values = annot_colors$ER1, guide = FALSE) +
        theme(aspect.ratio = 1) +
        xlim(0, 0.65) +
        ylim(0, 0.65) #+
#         annotate("text", x = 0.5, y = 0.1, label = paste("rho = ", round(cor(xist_meth_cna$`1N`, xist_meth_cna$`2N`, use = "pairwise.complete.obs"), digits = 3)), size = 2, family = "Arial")
(p_scatter_meth_cna_1_2 + theme_bw() + theme(aspect.ratio = 1)) + (p_scatter_meth_cna_2_3 + theme_bw() + theme(aspect.ratio = 1))
```

```{r}
p_scatter_meth_cna_2_3 + facet_wrap(~ER) + theme_bw() + theme(aspect.ratio = 1) 
```

Same with expression:

```{r}
get_xist_expr_cna
xist_expr_cna <- get_xist_expr_cna()
head(xist_expr_cna)
```

#### Extended Data Figure 9F

```{r}
p_scatter_expr_cna_2_3 <- xist_expr_cna %>%
        ggplot(aes(x = `2N`, y = `>=3N`, color = ER)) +
        geom_point(size=0.2) + 
        geom_abline(color = "black", linetype = "dashed") +
        scale_color_manual(values = annot_colors$ER1, guide = FALSE) +
        theme(aspect.ratio = 1) +
        xlim(5, 14) +
        ylim(5, 14)

 p_scatter_expr_cna_1_2 <- xist_expr_cna %>% ggplot(aes(x = `1N`, y = `2N`, color = ER)) +
        geom_point(size=0.2) + 
        geom_abline(color = "black", linetype = "dashed") +
        scale_color_manual(values = annot_colors$ER1, guide = FALSE) +
        theme(aspect.ratio = 1) +
        xlim(5, 14) +
        ylim(5, 14) 

p_scatter_expr_cna_1_2 + p_scatter_expr_cna_2_3
```

## autosome analysis

```{r}
get_autosome_loci
autosome_loci <- get_autosome_loci()
```

```{r}
get_autosome_cna
autosome_cna <- get_autosome_cna()
head(autosome_cna)
```

```{r}
get_autosome_loci_meth
autosome_loci_meth <- get_autosome_loci_meth()
head(autosome_loci_meth)
```

Plotting the distribution of loci


#### Extended Data Figure 9G

```{r}
options(repr.plot.width = 8, repr.plot.height = 5)
df <- autosome_cna %>%
    left_join(autosome_loci_meth) %>%
    group_by(cna_grp, ER, chrom, start, end) %>%
    summarise(meth = mean(meth, na.rm = TRUE)) %>%
    filter(!is.na(ER)) %>% 
    mutate(cna_grp = factor(cna_grp, levels = c("1N", "2N", ">=3N"))) %>% 
    mutate(ER = factor(ER, levels = c("ER+", "ER-")))
p_boxp_meth_cna_autosome <- df %>% 
    ggplot(aes(x = cna_grp, y = meth, fill = ER, group = cna_grp)) +
#     ggrastr::geom_boxplot_jitter(outlier.size = 0.1, outlier.jitter.width = 0.01, raster=TRUE) + 
    geom_boxplot(linewidth=0.1, fatten=0.5, outlier.size = 0.1) + 
    scale_fill_manual(values = annot_colors$ER1, guide = FALSE) +
    xlab("") +
    ylab("Methylation in autosomes") +
    facet_grid(. ~ ER) +
    ylim(0, 1.1) + 
            theme(
                panel.grid.major.x = element_blank(),
                panel.grid.minor.x = element_blank()
            )

p_boxp_meth_cna_autosome + ggpubr::stat_compare_means(comparisons = list(c("1N", "2N"), c("2N", ">=3N")))
p_boxp_meth_cna_autosome +
     ggpubr::stat_compare_means(label = "p.signif", comparisons = list(c("1N", "2N"), c("2N", ">=3N"))) 
p_boxp_meth_cna_autosome
```

```{r}
df %>% ungroup() %>% distinct(chrom, start, end, ER) %>% count(ER)
```

```{r}
autosome_cna %>% distinct(cna_grp, samp, ER) %>% count(ER, cna_grp)
```

We now move to show scatters of loci in different CNA regimes

```{r}
get_autosome_meth_cna
autosome_meth_cna <- get_autosome_meth_cna()
head(autosome_meth_cna)
```

#### Extended Data Figure 9H

```{r}
options(repr.plot.width = 5, repr.plot.height = 3)
p_scatter_meth_cna_2_3_autosome <- autosome_meth_cna %>%
        ggplot(aes(x = `2N`, y = `>=3N`, color = ER)) +
        geom_point(size=0.2) + 
        geom_abline(color = "black", linetype = "dashed") +
        scale_color_manual(values = annot_colors$ER1, guide = FALSE) +
        geom_abline(color = "red", linetype = "dashed", intercept = 0.1) + 
        geom_abline(color = "red", linetype = "dashed", intercept = -0.1) + 
        theme(aspect.ratio = 1)
p_scatter_meth_cna_1_2_autosome <- autosome_meth_cna %>%
        ggplot(aes(x = `1N`, y = `2N`, color = ER)) +
        geom_point(size=0.2) + 
        geom_abline(color = "black", linetype = "dashed") +
        scale_color_manual(values = annot_colors$ER1, guide = FALSE) +        
        geom_abline(color = "red", linetype = "dashed", intercept = 0.1) + 
        geom_abline(color = "red", linetype = "dashed", intercept = -0.1) +
        theme(aspect.ratio = 1)          
        
p_scatter_meth_cna_1_2_autosome + p_scatter_meth_cna_2_3_autosome
```

Same with expression:

```{r}
get_autosome_expr_cna
autosome_expr_cna <- get_autosome_expr_cna()
head(autosome_expr_cna)
```

```{r}
p_scatter_expr_cna_2_3 <- autosome_expr_cna %>%
        ggplot(aes(x = `2N`, y = `>=3N`, color = ER)) +
        geom_point(size=0.2) + 
        geom_abline(color = "black", linetype = "dashed") +
        scale_color_manual(values = annot_colors$ER1, guide = FALSE) +
        theme(aspect.ratio = 1) +
        xlim(5, 14) +
        ylim(5, 14)

 p_scatter_expr_cna_1_2 <- autosome_expr_cna %>% ggplot(aes(x = `1N`, y = `2N`, color = ER)) +
        geom_point(size=0.2) + 
        geom_abline(color = "black", linetype = "dashed") +
        scale_color_manual(values = annot_colors$ER1, guide = FALSE) +
        theme(aspect.ratio = 1) +
        xlim(5, 14) +
        ylim(5, 14) 

p_scatter_expr_cna_1_2 + p_scatter_expr_cna_2_3
```

Looking at the distribution of loci in 2N / >=3N (averaging the samples expression per ER and CNA)

```{r}
get_xist_expr_cna
get_autosome_expr_cna
df_fc <- get_xist_expr_cna() %>%
        mutate(type = "X") %>%
        bind_rows(get_autosome_expr_cna() %>% mutate(type = "auto")) %>%
        mutate(diff = `>=3N` - `2N`, diff1 = `2N` - `1N`) %>%
#         filter(ER == "ER+")
        filter(ER == "ER-")
head(df_fc)
```

#### Extended Data Figure 9I

```{r}
pval <- ks.test(df_fc$diff[df_fc$type == "X"], df_fc$diff[df_fc$type == "auto"])$p.value
message(pval)

fc_p <- df_fc %>%
        ggplot(aes(color = type, x = diff)) +
        geom_density(size = 1) +
        theme(aspect.ratio = 1) +
        xlab("log expression fold change (>=3N/2N)") +
        ylab("") +
        scale_color_manual(values = c("X" = "darkblue", "auto" = "red")) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        theme(aspect.ratio = 1) 
#         annotate("text", x = 0.5, y = 4, label = "p<<0.001 (KS)", size = 2, family = "Arial")

fc_p
```

```{r}
pval <- ks.test(df_fc$diff1[df_fc$type == "X"], df_fc$diff1[df_fc$type == "auto"])$p.value
message(pval)

fc_p <- df_fc %>%
        ggplot(aes(color = type, x = -diff1)) +
        geom_density(size = 1) +
        theme(aspect.ratio = 1) +
        xlab("log expression fold change (1N/2N)") +
        ylab("") +
        scale_color_manual(values = c("X" = "darkblue", "auto" = "red")) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        theme(aspect.ratio = 1) 
#         annotate("text", x = 0.5, y = 4, label = "p<<0.001 (KS)", size = 2, family = "Arial")

fc_p
```

```{r}
head(df_fc)
```

```{r}
thresh <- 0.1
autosome_cna_meth_expr <- get_autosome_cna_meth_expr()
    dosage_cands <- autosome_cna_meth_expr %>% 
        filter(`>=3N` >= (`2N` + thresh) ) %>%         
        filter(`n_>=3N` >= 5, `n_2N` >= 5) 

    dosage_cands_1N <- autosome_cna_meth_expr %>% 
        filter(`1N` <= (`2N` - thresh) ) %>%         
        filter(`n_1N` >= 5, `n_2N` >= 5)     

    dosage_cands_expr <- dosage_cands %>% select(chrom, start, end, ER, `1N` = `expr_1N`, `2N` = `expr_2N`, `>=3N` = `expr_>=3N`)
    dosage_cands_expr_1N <- dosage_cands_1N %>% select(chrom, start, end, ER, `1N` = `expr_1N`, `2N` = `expr_2N`)

    df_fc <- get_xist_expr_cna() %>%
        mutate(type = "X") %>%
        bind_rows(autosome_expr_cna %>% anti_join(dosage_cands_expr) %>% mutate(type = "auto")) %>%
        bind_rows(dosage_cands_expr %>% mutate(type = "auto-dosage")) %>% 
        mutate(ER = factor(ER, levels=c("ER+", "ER-"))) %>%
        filter(!is.na(ER)) %>% 
        mutate(type = factor(type, levels=c("X", "auto", "auto-dosage"))) %>%
        arrange(type) %>%
        mutate(diff = `>=3N` - `2N`) 

    fc_p <- df_fc %>%
        ggplot(aes(color = type, x = diff)) +
        geom_density(size = 0.5) +
        theme(aspect.ratio = 1) +
        xlab("log expression fold change (>=3N/2N)") +
        ylab("") +
        scale_color_manual(values = c("auto" = "red", "auto-dosage" = "blue", "X" = "gray")) + 
        # ggsci::scale_color_npg() +         
        coord_cartesian(xlim = c(-0.6, 1)) +
        theme(aspect.ratio = 1) + 
        facet_wrap(~ER, scale="free_y", nrow=1) + 
        guides(color = FALSE)

    df_fc %>% group_by(ER) %>% summarise(p_x_auto = ks.test(diff[type == "X"], diff[type == "auto"])$p.value, p_auto_dosage =  ks.test(diff[type == "auto-dosage"], diff[type == "auto"])$p.value) %>% print()

    df_fc_1N <- get_xist_expr_cna() %>%
        mutate(type = "X") %>%
        bind_rows(autosome_expr_cna %>% anti_join(dosage_cands_expr_1N) %>% mutate(type = "auto")) %>%
        bind_rows(dosage_cands_expr_1N %>% mutate(type = "auto-dosage")) %>% 
        mutate(ER = factor(ER, levels=c("ER+", "ER-"))) %>%
        filter(!is.na(ER)) %>% 
        mutate(type = factor(type, levels=c("X", "auto", "auto-dosage"))) %>%
        arrange(type) %>%
        mutate(diff = `1N` - `2N`) 

    fc_p_1N <- df_fc_1N %>%
        ggplot(aes(color = type, x = diff)) +
        geom_density(size = 0.5) +
        theme(aspect.ratio = 1) +
        xlab("log expression fold change (1N/2N)") +
        ylab("") +
        scale_color_manual(values = c("auto" = "red", "auto-dosage" = "blue", "X" = "gray")) +           
        coord_cartesian(xlim = c(-0.6, 1)) +
        theme(aspect.ratio = 1) + 
        facet_wrap(~ER, scale="free_y", nrow=1) + 
        guides(color = FALSE)

    df_fc_1N %>% group_by(ER) %>% summarise(p_x_auto = ks.test(diff[type == "X"], diff[type == "auto"])$p.value, p_auto_dosage =  ks.test(diff[type == "auto-dosage"], diff[type == "auto"])$p.value) %>% print() 

fc_p
fc_p_1N
```

## Checking dosage compensation on all X loci

```{r}
get_xist_loci
```

```{r}
all_norm_meth_x <- fread(here("data/all_norm_meth.tsv")) %>% 
            filter(chrom == "chrX") %>% as_tibble()
X_loci <- all_norm_meth_x %>% distinct(chrom, start, end)
```

```{r}
get_xist_cna
```

```{r}
X_cna <- X_loci %>% gintervals.neighbors1(cna %>% 
    mutate(end = ifelse(start == end, start + 1, end)) %>% 
    filter(chrom == "chrX"), maxneighbors = nrow(samp_data)) %>% 
    filter(dist == 0) %>% select(chrom, start, end, samp, 
    cna = cna_round)
X_cna <- X_cna %>% mutate(cna_grp = cut(cna, breaks = c(-1, 
        0, 1, 2, 10), labels = c("0N", "1N", "2N", ">=3N"))) %>% 
        filter(cna_grp != "0N") %>% left_join(samp_data %>% select(samp, 
        ER = ER1)) %cache_df% here("data/X_cna.tsv") %>% as_tibble()
head(X_cna)
```

```{r}
get_xist_loci_meth
```

```{r}
X_loci_meth <- X_loci %>% inner_join(get_promoter_avg_meth()) %>% 
        gather("samp", "meth", -(chrom:end))
```

```{r}
options(repr.plot.width = 5, repr.plot.height = 3)
p_boxp_meth_cna_all <- X_cna %>%
    left_join(X_loci_meth) %>%
    group_by(cna_grp, ER, chrom, start, end) %>%
    summarise(meth = mean(meth, na.rm = TRUE)) %>%
    filter(!is.na(ER)) %>% 
    mutate(cna_grp = factor(cna_grp, levels = c("1N", "2N", ">=3N"))) %>% 
    mutate(ER = factor(ER, levels = c("ER+", "ER-"))) %>% 
    ggplot(aes(x = cna_grp, y = meth, fill = ER, group = cna_grp)) +
    geom_boxplot(linewidth=0.1, fatten=0.5, outlier.size = 0.1) + 
    scale_fill_manual(values = annot_colors$ER1, guide = FALSE) +
    xlab("") +
    ylab("Methylation in X promoters") +
    facet_grid(. ~ ER) +
    ylim(0, 1.1) +
    ggpubr::stat_compare_means(label = "p.signif", comparisons = list(c("1N", "2N"), c("2N", ">=3N"))) + 
            theme(
                panel.grid.major.x = element_blank(),
                panel.grid.minor.x = element_blank()
            )
p_boxp_meth_cna_all
```

> Looks very similar to the XIST associated version


## % of X lost


#### Extended Data Figure 9C

```{r}
chr_x_len <- gintervals.all() %>% filter(chrom == "chrX") %>% pull(end)
p_x_perc <- cna %>% 
    filter(chrom == "chrX") %>% mutate(l = end - start) %>% 
    mutate(cna = cut(cna_round, breaks = c(0, 1, 2, 20), include.lowest=TRUE, labels=c("Loss", "normal", "Gain"))) %>% 
    group_by(samp, cna) %>% 
    summarise(p = sum(l) / chr_x_len) %>% 
    filter(cna != "normal") %>% 
    ungroup() %>% 
    add_ER() %>% 
    filter(!is.na(ER)) %>% 
    ggplot(aes(x=p, color=ER, y=1-..y..)) + 
        stat_ecdf() + 
        facet_wrap(~cna) + 
        scale_color_manual(values=annot_colors$ER1) + 
        theme(aspect.ratio=1) + 
        xlab("% of X") + 
        ylab("% of samples") + 
        scale_x_continuous(labels=scales::percent) + 
        scale_y_continuous(labels=scales::percent)
p_x_perc
```

## Plot Expression-Methylation correlation for X

```{r}
meth_mat <- all_norm_meth_x %>% select(chrom:end, any_of(ER_positive_samples)) %>% intervs_to_mat()
```

```{r}
expr_m <- get_gene_expression_mat() %>% select(-any_of(c("chrom", "start", "end", "name3.chr")))
expr_mat <- expr_m %>%
    as.data.frame() %>%
    column_to_rownames("name")
f <- rowSums(!is.na(expr_mat)) > 0
expr_mat <- expr_mat[f, ]
dim(expr_mat)
```

```{r}
em_cross <- em_cross_cor(meth_mat, expr_mat, meth_cor_thresh = 0.25, expr_cor_thresh = 0.25) %cache_rds% here("data/X_er_positive_em_cross_cor.rds")
```

```{r}
em_cross_clust <- cluster_em_cross_cor(em_cross, k_meth = 32, k_expr = 32) %cache_rds% here("data/X_er_positive_em_cross_cor_clust.rds")
```

#### Extended Data Figure 9A

```{r}
options(repr.plot.width = 8, repr.plot.height = 13)
plot_em_cross_cor(em_cross_clust)
```

```{r}
em_cross_clust$em_cross[1:5, 1:5]
```

```{r}
dim(em_cross_clust$em_cross)
```

# Finding specific dosage compensation examples

```{r}
autosome_cands_3n <- get_autosome_dosage_comp_cands("3N")
head(autosome_cands_3n)
```

```{r}
thresh <- 0.1
dosage_cands_gain_3n <- autosome_cands_3n %>% 
        filter(`>=3N` >= (`2N` + thresh) ) %>%         
        filter(`n_>=3N` >= 5, `n_2N` >= 5) %>% #,  !is.na(name))  %>%
        select(chrom, start, end, name, ER, `>=3N`, `2N`, `n_>=3N`, n_2N, `expr_>=3N`, `expr_2N`, `n_expr_>=3N`, `n_expr_2N`) %>%
        arrange(-`>=3N`, ER, name) 
dosage_cands_gain_3n %>% count(ER)
```

```{r}
autosome_cands_4n <- get_autosome_dosage_comp_cands("4N")
head(autosome_cands_4n)
```

```{r}
thresh <- 0.1
dosage_cands_gain_4n <- autosome_cands_4n %>% 
        filter(`>=4N` >= (`2N` + thresh) ) %>%         
        filter(`n_>=4N` >= 5, `n_2N` >= 5) %>% #,  !is.na(name))  %>%
        select(chrom, start, end, name, ER, `>=4N`, `2N`, `n_>=4N`, n_2N, `expr_>=4N`, `expr_2N`, `n_expr_>=4N`, `n_expr_2N`) %>%
        arrange(-`>=4N`, ER, name) 
dosage_cands_gain_4n %>% count(ER)
```

```{r}
dosage_cands_loss <- autosome_cna_meth_expr %>% 
        filter(`1N` <= (`2N` - thresh) ) %>%         
        filter(`n_1N` >= 5, `n_2N` >= 5)  %>%
        select(chrom, start, end, name, ER, `1N`, `2N`, `n_1N`, n_2N, `expr_1N`, `expr_2N`, `n_expr_1N`, `n_expr_2N`) %>%
        arrange(-`n_1N`, ER, name)

dosage_cands_loss %>% count(ER)
```

```{r}
df <- get_autosome_meth_cna()
```

Validating that we have enough samples for each locus:

```{r}
auto_cna <- get_autosome_cna() %>% inner_join(get_autosome_loci_meth())
```

```{r}
num_cna_meth <- auto_cna %>% filter(!is.na(meth)) %>% count(chrom, start, end, ER, cna_grp) %>% mutate(cna_grp = paste0("n_", cna_grp)) %>% spread(cna_grp, n)
```

```{r}
num_cna_meth %>% gather('type', 'num', -(chrom:ER)) %>% group_by(chrom, start, end, ER) %>% mutate(p = num / sum(num))  %>% select(-num) %>% spread(type, p) %>% ggplot(aes(x=`n_1N`, y=`n_2N`, color=ER)) + geom_point() + scale_color_manual(values=annot_colors$ER1, guide = FALSE) + theme(aspect.ratio=1)
num_cna_meth %>% gather('type', 'num', -(chrom:ER)) %>% group_by(chrom, start, end, ER) %>% mutate(p = num / sum(num))  %>% select(-num) %>% spread(type, p) %>% ggplot(aes(x=`n_2N`, y=`n_>=3N`, color=ER)) + geom_point() + scale_color_manual(values=annot_colors$ER1, guide = FALSE) + theme(aspect.ratio=1)
```

```{r}
df <- df %>% left_join(num_cna_meth)
```

```{r}
df_expr <- get_autosome_expr_cna() %>% gather("cna", "expr", -(chrom:ER)) %>% mutate(cna = paste0("expr_", cna)) %>% spread(cna, expr)
```

```{r}
df <- df %>% left_join(df_expr)
```

```{r}
num_cna_expr <- get_autosome_cna() %>% inner_join(get_autosome_loci_expr()) %>% filter(!is.na(expr)) %>% count(chrom, start, end, ER, cna_grp) %>% mutate(cna_grp = paste0("n_expr_", cna_grp)) %>% spread(cna_grp, n)
```

```{r}
df <- df %>% left_join(num_cna_expr)
```

```{r}
df <- df %>% left_join(get_gene_expression_mat() %>% select(chrom:end, name) %>% mutate(start = start + 1, end = end + 1))
```

```{r}
fwrite(df, here("data/autosome_cna_meth_expr.tsv"), sep="\t")
```

```{r}
thresh <- 0.1
```

```{r}
options(repr.plot.width = 4, repr.plot.height = 4)

df %>% ggplot(aes(x = `2N`, y = `>=3N`, color = ER)) +geom_point(size=0.2)  + geom_abline(color = "black", linetype = "dashed") + scale_color_manual(values = annot_colors$ER1, guide = FALSE) + theme(aspect.ratio = 1) + geom_line(data = tibble(`2N` = seq(0,1,0.001), `>=3N`= `2N` - thresh), color="red", linetype="dashed") +  geom_line(data = tibble(`2N` = seq(0,1,0.001), `>=3N`= `2N` + thresh), color="red", linetype = "dashed") + xlim(0,1) + ylim(0,1)

```

```{r}
df %>% filter(`>=3N` >= (`2N` + thresh) ) %>% select(chrom, start, end, ER, `>=3N`, `2N`, `n_>=3N`, n_2N, `expr_>=3N`, `expr_2N`, `n_expr_>=3N`, `n_expr_2N`, name) %>% filter(`n_>=3N` >= 5, `n_2N` >= 5, !is.na(name)) %>% arrange(-`n_>=3N`, ER, name) %>% ggplot(aes(x=(`expr_>=3N` - `expr_2N`))) + geom_density() + xlim(-0.5, 1)
```

```{r}
df %>% filter(`>=3N` >= (`2N` + thresh) ) %>% select(chrom, start, end, ER, `>=3N`, `2N`, `n_>=3N`, n_2N, `expr_>=3N`, `expr_2N`, `n_expr_>=3N`, `n_expr_2N`, name) %>% filter(`n_>=3N` >= 5, `n_2N` >= 5, !is.na(name)) %>% arrange(-`n_>=3N`, ER, name)
```

```{r}
df %>% filter(`>=3N` <= (`2N` - thresh) ) %>% select(chrom, start, end, ER, `>=3N`, `2N`, `n_>=3N`, n_2N, `expr_>=3N`, `expr_2N`, `n_expr_>=3N`, `n_expr_2N`, name) %>% filter(`n_>=3N` >= 5, `n_2N` >= 5, !is.na(name))
```

```{r}
dosage_cands <- df %>% filter(`>=3N` >= (`2N` + thresh) ) %>% select(chrom, start, end, ER, `>=3N`, `2N`, `n_>=3N`, n_2N, `expr_>=3N`, `expr_2N`, `n_expr_>=3N`, `n_expr_2N`, name) %>% filter(`n_>=3N` >= 5, `n_2N` >= 5, !is.na(name)) %>% arrange(-`n_>=3N`, ER, name)
```

```{r}
dosage_cands_mean_expr <- dosage_cands %>% select(chrom, start, end, ER) %>% inner_join(get_gene_expression_mat() %>% select(-name3.chr) %>% mutate(start = start + 1, end = end + 1))  %>% gather("samp", "expr", -(chrom:name)) %>% left_join(samp_data %>% select(samp, ER1)) %>% filter(ER == ER1) %>% group_by(chrom, start, end, name, ER) %>% summarise(mean_expr = mean(expr, na.rm=TRUE)) %>% ungroup()
all_mean_expr <- get_gene_expression_mat() %>% select(-name3.chr) %>% mutate(start = start + 1, end = end + 1)  %>% gather("samp", "expr", -(chrom:name)) %>% left_join(samp_data %>% select(samp, ER1)) %>% group_by(chrom, start, end, name, ER1) %>% summarise(mean_expr = mean(expr, na.rm=TRUE)) %>% ungroup()
```

```{r}
dosage_cands_mean_expr %>% mutate(type = "dosage") %>% bind_rows(all_mean_expr %>% mutate(type = "all")) %>% ggplot(aes(x=mean_expr, color=type)) + geom_density()
```

```{r}
get_xist_cna() %>%
        left_join(get_xist_loci_expr()) %>%
        group_by(cna_grp, samp, ER, chrom, start, end) %>%
        summarise(expr = mean(expr, na.rm = TRUE)) %>%
        ggplot(aes(x = cna_grp, y = expr, fill = ER, group = cna_grp)) +
        geom_boxplot(outlier.size = 0.1) + 
        scale_fill_manual(values = annot_colors$ER1, guide = FALSE) +
        xlab("") +
        ylab("Expression in Xist\nassociated promoters") +
        facet_grid(. ~ ER) +        
        ggpubr::stat_compare_means(label = "p.signif", comparisons = list(c("1N", "2N"), c("2N", ">=3N")))
```

```{r}
gc()
```
