---
title: "METABRIC-RRBS pipeline"
output:
  html_document:
    toc: true
    toc_float: true   
    number_sections: false
    theme: united
    highlight: tango
    # code_folding: show 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Workdir and other definitions

### Load packages

```{r}
library(tidyverse)
library(glue)
library(gpatterns)
library(tgutil)
```

### Definitions
Set the working directory (change the code below if you are not running from the base repository): 

```{r}
workdir <- getwd()
```

Set the Misha database: 

```{r}
gsetroot('db/trackdb')
```

# Annotating the BAM file names

We list all the bam files and parse the file names:

```{r}
bams <- tibble(
        bam = list.files(
            glue("{workdir}/rawdata/BAM"), 
            include.dirs=FALSE, 
            full.names=TRUE, 
            recursive=TRUE, 
            pattern="*\\.bam$")
        ) %>% 
    mutate(fn = basename(bam), 
           fn = gsub(".fq.gz_bismark_bt2.bam", "", fn), 
           bam_dir = dirname(bam), 
           tcpgs_dir = gsub("BAM", "TCPGS", bam_dir)) %>%
    mutate(
        batch = 
            gsub(glue("{workdir}/rawdata/BAM/"), "", bam) %>% 
            gsub("/.*", "", .), name=gsub(".s_..r_1_trimmed", "", fn), 
            name = paste0(batch, '.', name), 
            name=gsub('-', '_', name), 
        track=paste0('metabric_rrbs.', name)
        ) %>%
    mutate(tcpgs_dir = paste0(tcpgs_dir, "/", track)) %>%
    select(bam, tcpgs_dir, track)
```

# Running the pipeline

We generate a command for each bam file (sample):

```{r}
cmds <- bams %>% 
    mutate(cmd = glue('{{ \\
        gpatterns::gpatterns.set_parallel(8); \\
        gpatterns::gpatterns.import_from_bam(\\
        "{bam}", \\
        workdir=NULL, \\
        steps=c("bam2tidy_cpgs", "bind_tidy_cpgs", "pileup", "pat_freq"), \\
        use_sge=F, \\
        parallel=TRUE, \\
        bismark=TRUE, \\
        add_chr_prefix=TRUE, \\
        paired_end=FALSE, \\
        track="{track}", \\
        description="xxx", \\
        pat_freq_len=5, \\
        min_qual=20)}}')) %>% 
    pull(cmd)
```

Run the pipeline: 

```{r}
message(glue('glue running {length(cmds)} jobs'))
res <- gcluster.run2(command_list = cmds, jobs_title="metabric_rrbs", max.jobs=25, threads=5, io_saturation=1)
```

### Make sure that all the samples succeeded
Update misha database: 

```{r}
gdb.reload()
```

Test that samples have all the tidy_cpgs files and misha tracks

```{r}
get_chrom_stats <- function(track){
    tcpgs_dir <- paste0(gpatterns:::.gpatterns.base_dir(track), '/tidy_cpgs')

    df <- tibble(fn = list.files(tcpgs_dir, full.names=T)) %>% 
        mutate(size = file.size(fn)) %>%
        filter(grepl('chr', basename(fn))) %>% 
        mutate(chrom = gsub('\\.tcpgs\\.gz$', '', basename(fn))) %>% 
        mutate(stat_fn = paste0(tcpgs_dir, '/stats/', chrom, '.stats')) %>% 
        filter(file.exists(stat_fn))
    df <- df %>% plyr::adply(1, function(x) fread(x$stat_fn)) %>% as_tibble() %>% mutate(track = track)
    return(df)
}

stats <- bams %>% plyr::adply(1, function(x) get_chrom_stats(x$track), .parallel=TRUE) %>% as_tibble()
stats <- stats %>% 
    mutate(track_exists = gtrack.exists(paste0(track, '.cov')) & gtrack.exists(paste0(track, '.meth')) & gtrack.exists(paste0(track, '.unmeth')))
stats <- stats %>% 
    mutate(pat_track = paste0(track, '.pat5'), pat_track_exists=gtrack.exists(pat_track))

failed_samples <- stats %>% filter(!track_exists | !pat_track_exists | (size < 100 & !grepl('chrY', chrom))) %>% distinct(bam, track)
failed_samples

```


# Sample QC

```{r}
get_track_stats <- function(track){
    sm <- gsummary(glue('{track}.cov'))
    stats <- tibble(
        cg_num = sm[1],
        meth_calls = sm[5],
        global_avg_meth = gsummary(glue("{track}.avg"))[6]
        )
    return(stats)
}

get_tracks_qc <- function(){
    track_stats <- bams %>% 
        plyr::adply(1, function(x) get_chrom_stats(x$track), .parallel=TRUE) %>% 
        group_by(bam, track) %>% 
        summarise(
            CpG = mean(CpG, na.rm=TRUE), 
            CHH = mean(CHH, na.rm=TRUE), 
            CHG = mean(CHG, na.rm=TRUE), 
            single_R1 = single_R1[1], 
            total_reads = single_R1 + no_conv[1] + unmapped[1]
            ) %>%
        mutate(mapped_reads = single_R1, mapped_frac = mapped_reads / total_reads) %>% 
        select(-single_R1) %>%
        ungroup() %>%
        plyr::adply(1, function(x) { message(x$track); get_track_stats(x$track) }, .parallel=TRUE) %>%
        as_tibble()
    return(track_stats)
}

qc <- get_tracks_qc()

fwrite(qc, 'data/sample_qc.csv')

qc
```